Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-14 FGCS_rebuttal.txt

We value the effort invested by the reviewers in assessing our paper.
A list of changes for the raised points:
-------------
* [Introduction clearly describes one of the main interests of the paper: a methodology for building models (in the model-checking sense) for a legacy system. On the other hands, it fails to introduce the second contribution of the paper, which is a methodology to actually performing the check...]

> Introduction has been modified to introduce these aspects, paragraph:
[Once the model is in place, the toolset can automatically check for general system properties, such as deadlocks. We show how to formulate and check application-specific properties... 
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent systems.]
-------------
* [However, for a reader not familiar with the history of the EGEE/EGI/WLCG grid, some information is missing. Mainly, Dirac appears as a standalone system, and the underlying software and hardware infrastructure is not described.
- The paper should explain the relation of Dirac with the gLite middleware, and provide a reference to  glite (eg E. Laure, S.M. Fisher, A. Frohner, C. Grandi, P. Kunszt, et al. Programming the Grid with gLite. Computational Methods in Science and Technology, 12(1):33-45, 2006.).]

> This explanation has been added in Section 2.2, paragraph:
[Originally, the DIRAC WMS submitted pilot jobs using the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers... making sites compete with each other for jobs and have a faster job turnaround.]
> gLite reference also added
-------------
* [The paper should give figures showing the size and scale of distribution of the hardware infrastructure (number of sites, of CPUs or cores, storage capacity etc.) accessible to DIRAC.]
> Paragraph added:
[It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over tens of grid sites, ....About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps.] 
> Figures 1 and 2 added for illustration.
-------------
* [Considering the SMS, a state  machine analogous to the one for WMS would improve the readability of section 2.3.]
> Figure 7 added.
-------------
* [ Other approaches of model checking (eg deductive verification) could have been discussed. Also, the choice of model checking versus trace mining (a very active research area) should be justified, as both Dirac and gLite services generate copious logs that were not exploited for diagnosis. Overall, a Related Work section is necessary.]
> Related work was addressed by all reviewers, so we dedicated a separate Section 5 for this.

* [Section 4 is also very pedagogical: it follows the real-world checking process, including preliminary visualization. However, figures 7 should be provided with a better resolution and better explained ... Figure 8 should probably printed in color, or entirely removed. If kept, the "suspicious transition" should be pointed in the figure.]
> Figure 7 (now 10) is now with a better resolution (background circles simply indicate which portion of the entire state space belongs to this self-transitions. We found it un-necessary to bother the reader with this detail). Figure 8 (now 11): suspicious transition is clearly pointed at, now.
-------------
* [Section 4.3 demonstrates the efficiency of the approach, on the smallest of the two state machine. On the other hand, the approach fails on the WMS, due to combinatorial explosion.  The strategy to cope with this issue is presented as a "trick", and should be better related to the relevant concepts in the literature, in particular to testing.]
> We briefly relate this in Section 5. Related Work [This component resembles our monitoring process automaton used to cope with state space explosion during model checking of DIRAC’s WMS system]
-------------
* [The authors motivate the problem of lost efficiency of resource usage and manpower. Are there any statistics showing actual loss of efficiency? How often the DIRAC system get into inconsistent state?]
> Added in Introduction: [Depending on the frequency of certain heavier data (re)processing campaigns throughout the year, the occurrence of such inconsistencies can be up to several reported cases per month..., compared to a temporary fix, like manipulating database records or restarting processes manually]
-------------
* [The application of model checking on DIRAC system and finding race conditions is useful. However, the extendability of this approach for other systems is not clear to me. Can these approaches be applied on other systems? What changes need to be done?]
> We attempt to tackle this in Section 6. Conclusions. A few more lines of explanation are added in: [It is observed that many large concurrent/distributed systems are essentially parametrized, meaning that they contain components or process descriptions which are behaviorally similar...and we believe that the combination of code slicing and abstraction rules of thumb presented in Section 3 can be helpful in this direction.]
-------------
* Reviewer #4's comments again raise the issues of applicability of the approach to other systems, and in depth discussion of related work. (covered above)
-------------
* [Some discussion on the complexity and/or time requirements of using the model would be interesting to understand the usefulness of the model]
> This is difficult to generalize for any model, but we provide statistics on the state space generation and model checking times in section 4.Analysis and Issues.

Best regards,
Daniela

### 2013-02-14 ElsevierFGCS_singleColumn.tex
We're almost ready to go.

Although much effort is invested in making DIRAC reliable, entities occasionally
get into inconsistent states, leading to a loss of efficiency in both
resource usage and manpower. Debugging and fixing the root of such encountered
behaviors becomes a formidable mission due to multiple factors: the inherent
parallelism present among the system components deployed on different physical
machines, the size of the implementation ($\sim$150,000 lines of Python code),
and the distributed knowledge of different subsystems within the collaboration.
Depending on the frequency of certain heavier data (re)processing campaigns throughout 
the year, the occurrence of such inconsistencies can be up to several reported cases 
per month. However, since they have a blocking effect on these heavy grid activities
with a strict deadline, locating and fixing the root cause of such a problem 
typically has a lower priority within the collaboration, compared to a temporary fix, 
such as manipulating database records or restarting processes manually.

In this paper we propose the use of rigorous (formal) methods for improving
software quality. Model checking \cite{ProcessesWithData} is one such technique for analysis of an
abstract model of a system, and verification of certain system properties of
interest. Unlike conventional testing, it allows full control over the execution
of parallel processes and also supports automated exhaustive state-space
exploration. We used the mCRL2 language \cite{FormalLanguagemCRL2} and toolset \cite{mCRL2Toolset} to model the behavior of two
critical and related DIRAC components: the Workload Management (WMS) and the
Storage Management System (SMS). Based on Algebra of Communicating Processes
(ACP) \cite{process_algebra}, mCRL2 is able to deal with generic data types as well as user-defined
functions for data transformation. This makes it particularly suitable for
modeling the data manipulations made by DIRAC's agents. Visualizing the state
space and replaying scenarios with the toolkit's simulator enabled us to gain
insight into the system behavior and incrementally improve the model. 
Already using these techniques, critical race conditions and livelocks were detected and confirmed to
occur in the real system. Some of them were a result of simple coding bugs;
others unveiled more elementary design problems. We further formulated,
formalized and verified several application-specific properties. Once
the model is in place, the toolset can automatically check for general system
properties, such as deadlocks. We show how to formulate and check
application-specific
properties with the toolset. Furthermore, we outline a methodology to
effectively tackle the classical state space explosion
problem arising when the underlying model grows beyond the resources available
to perform explicit model checking. For this, we borrow 
concepts from finite state machine formalisms for description of system properties.
Extending the specification with a so-called monitoring automaton for tracking 
relevant transitions in the original model, can speed up the
analysis and discovery of traces that lead to violation of a certain property.
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent programs.

@@

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs
(Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day,
during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily
data transfers between storage units reaching 3Gbps
(Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}

@@

\textit{Agents} are active components that fulfill a limited number of specific system
functions. They run in different environments, depending on their mission.
Some are deployed close to the corresponding services, while others run on the
grid worker nodes. Examples of the latter are the so-called Pilot Agents, part
of the Workload Management System explained in the following section. All DIRAC
agents repeat the same logic in each iteration cycle: they monitor the service states,
 and respond by initiating actions (like job
submission or data transfer) which update the states of various system
entities.

\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

@@

The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' replica entries, checks whether they are registered in a Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the state of the \textit{CacheReplicas} and the related \textit{Tasks}
entries to ``Failed''. Non-problematic files are updated to a
``Waiting'' state. The \textit{Stage Request Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by Storage Element (SE) prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable, or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{Stage Monitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the \textit{CacheReplicas} and their
corresponding \textit{Tasks} entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{Request Finalization Agent}. The \textit{Tasks} which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated \textit{Tasks} for particular replicas, the respective
\textit{CacheReplicas} entries are also removed. The state machine of the \textit{CacheReplicas}
entity is shown in Fig.~\ref{fig:CacheReplicas-SM}.

@@

\subsubsection{Control Abstractions}
As already explained, all agents repeat the same logic in every subsequent iteration: first they
read some entries of interest from the service database, then they process the cached data, and finally
they may write back or update entries, based on decisions from the processing step.
They can be naturally modeled as \textit{recursive processes}. Each process represents a distinct agent,
specifying its expected behavior when interacting with the service database.
Take as an example the code snippet
in Listing~\ref{listing2}, from the \textit{Request Preparation Agent}. 
Although much of the code is omitted for clarity, the necessary parts for
illustrating the basic idea are kept.
The first highlighted statement is the selection of all "New'' CacheReplicas
entries. What follows is retrieval of their metadata from the LFC,
external to this subsystem. Subsequently, list and dictionary manipulations
are done to group the retrieved data depending on the outcome.
Two lists of replica IDs are built before the last step: one for the problematic catalog entries,
and one for the successful sanity checks.
Finally, the last two highlighted code segments update the states 
of the corresponding CacheReplicas to "Failed'' and ``Waiting'' respectively.
\definecolor{light-gray}{gray}{0.90}
\lstset{language=Python,tabsize=2,caption={\textit{RequestPreparationAgent.py} code excerpt},
  stringstyle=\ttfamily,
  %basicstyle=\small,
  frame=tb,
  breaklines=true,
  showstringspaces=false,
  boxpos=t
 }
\begin{lstlisting}[label=listing2,escapechar=!,basicstyle=\ttfamily\fontsize{8}{7}\ \selectfont]
  def prepareNewReplicas( self ):
    !\colorbox{light-gray}{res = self.getNewReplicas()}!
    if not res['Value']:
      gLogger.info('There were no New replicas found')
      return res
    [...]
    # Obtain the replicas from the FileCatalog
    res = self.__getFileReplicas( fileSizes.keys() )
    if not res['OK']:
      return res
    failed.update( res['Value']['Failed'] )
    terminal = res['Value']['ZeroReplicas']
    fileReplicas = res['Value']['Replicas']
    [...]
    replicaMetadata = []
    for lfn, requestedSEs in replicas.items():
      lfnReplicas = fileReplicas[lfn]
      for requestedSE, replicaID in requestedSEs.items():
        if not requestedSE in lfnReplicas.keys():
          terminalReplicaIDs[replicaID] = "LFN not registered at requested SE"
          replicas[lfn].pop(requestedSE)
        else:
          replicaMetadata.append((replicaID, lfnReplicas[requestedSE], fileSizes[lfn]))

@@

@@

\begin{figure*}[bp!]

@@

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478

@@

To overcome the problem of state space explosion, several
variants to explicit model checking have gained popularity. Runtime
verification~\cite{DBLP:conf/dagstuhl/ColinM04,Havelund:2008:RVC:1424465.1424470}
aims to increase confidence in the correctness, but does not make claims for
absence of defects in a system.
The target program is instrumented manually, or automatically by scanning
function calls and variable assignments, 
in order to capture and emit relevant events which are then monitored against a
special component during execution.
This component resembles our monitoring process automaton used to cope with
state space explosion during model checking 
of DIRAC's WMS system. While runtime verification is useful for
partially-capturing incorrect behavior of the system implementation, 
code instrumentation is not always a feasible option, considering the increase
of memory consumption.

@@

Static analysis is, in a sense, an extreme form of automated deductive
verification.  In its original form, deductive verification technology
allowed for the use of logics such as Floyd-Hoare logic~\cite{Hoa:69} for
showing the correctness of a software or hardware system by constructing
proofs thereof, based on mathematical descriptions of the behavior of
the system.  These days, the proofs involved are often discharged using a
mechanical or an automated theorem prover. Moreover, more advanced logics
such as Separation Logic~\cite{Rey:02} are used. Deductive verification
is applied quite successfully on software code~\cite{PMSJP:12,KAE+:10}
and hardware~\cite{AATH:10}, but the effort required for conducting the
verification can be prohibiting, and can compare poorly to model checking.

@@

Practitioners
of model checking have already built sufficiently mature tools
that can be utilized (almost) as blackboxes by more traditional
software developers, hiding the mathematical theories under
the hood. Although some domain-specific knowledge is always necessary,
the descriptive guidelines we have proposed for obtaining
an abstract model can be easily 
applied to other large-scale industrial systems of a similar nature.
It goes without saying that the analyst must be careful to faithfully
describe the system behavior in the abstract model. With this critical step in mind, formal methods
are an important addition to the assembly of software analysis and verification techniques for concurrent distributed systems.
\section{Future work}
\label{sec:Section_6}
Besides discovering behavioral problems, the 
model gives opportunities to analyze the system performance in the future.
Performance analysis aims at predicting the behavior of a system with respect to dynamic properties, such as 
the expected throughput, response time, resource usage, number of requests per unit time, etc.
For this, the formal model must be enriched with timing and probability quantities attached 
to certain activities of interest.  Adding this kind of information to traditional process algebra models
means getting in the domain of stochastic process algebra. Random variables with exponential distribution are
typically associated with the designated actions, to express system transitions with duration, which could be estimates
of processing time or communication overhead. Tools based on stochastic process algebra formalisms can
then generate the underlying Markov Chain model representation and solve it to obtain steady-state probability distributions over the state space. These measures can provide a good base for assessing the system's efficiency and reliability. 
Furthermore, by varying the quantitative inputs, alternative designs can be evaluated before actual implementation
takes place. One such stochastic process algebra is PEPA (Performance Evaluation Process Algebra \cite{PEPA}), supported
by an Eclipse Plug-in for modeling and analysis of systems. 
PEPA models are annotated with information about the duration of activities and their relative probabilities of occurence.

>>>>

Although much effort is invested in making DIRAC reliable, entities occasionally
get into inconsistent states, leading to a loss of efficiency in both
resource usage and manpower. Debugging and fixing the root of such encountered
behaviors becomes a formidable mission due to multiple factors: the inherent
parallelism present among the system components deployed on different physical
machines, the size of the implementation ($\sim$150,000 lines of Python code),
and the distributed knowledge of different subsystems within the collaboration.
Depending on the frequency of certain heavier data (re)processing campaigns throughout 
the year, the occurrence of such inconsistencies can be up to several reported cases 
per month. However, since they have a blocking effect on these heavy grid activities
with a strict deadline, locating and fixing the root cause of such a problem 
typically has a lower priority within the collaboration, compared to a temporary fix, 
like manipulating database records or restarting processes manually.

In this paper we propose the use of rigorous (formal) methods for improving
software quality. Model checking \cite{ProcessesWithData} is one such technique for analysis of an
abstract model of a system, and verification of certain system properties of
interest. Unlike conventional testing, it allows full control over the execution
of parallel processes and also supports automated exhaustive state-space
exploration. We used the mCRL2 language \cite{FormalLanguagemCRL2} and toolset \cite{mCRL2Toolset} to model the behavior of two
critical and related DIRAC components: the Workload Management (WMS) and the
Storage Management System (SMS). Based on Algebra of Communicating Processes
(ACP) \cite{process_algebra}, mCRL2 is able to deal with generic data types as well as user-defined
functions for data transformation. This makes it particularly suitable for
modeling the data manipulations made by DIRAC's agents. Visualizing the state
space and replaying scenarios with the toolkit's simulator enabled us to gain
insight into the system behavior and incrementally improve the model. 
Already using these techniques, critical race conditions and livelocks were detected and confirmed to
occur in the real system. Some of them were a result of simple coding bugs;
others unveiled more elementary design problems. We further formulated,
formalized and verified several application-specific properties. Once
the model is in place, the toolset can automatically check for general system
properties, such as deadlocks. We show how to formulate and check
application-specific
properties with the toolset. Furthermore, we outline a methodology to
effectively tackle the classical state space explosion
problem arising when the underlying model grows beyond the resources available
to perform explicit model checking. For this, we borrow 
concepts from finite state machine formalisms for description of the system properties.
Extending the specification with a so-called monitoring automaton for tracking 
relevant transitions in the original model, can speed up the
analysis and discovery of traces that lead to violation of a certain property.
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent systems.

@@

\begin{figure}[tp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure1_1.png}
\centering
\caption{LHCb job troughput by site}
\label{fig:Job-throughput}
\end{figure}
\begin{figure}[tp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure1_2.png}
\centering
\caption{LHCb data transfers throughput}
\label{fig:Data-transfers}
\end{figure}
The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs
(Fig.~\ref{fig:Job-throughput}) and over 100K jobs executed per day,
during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily
data transfers between storage units reaching 3Gbps
(Fig.~\ref{fig:Data-transfers}).

Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

@@

\textit{Agents} are active components that fulfill a limited number of specific system
functions. They run in different environments, depending on their mission.
Some are deployed close to the corresponding services, while others run on the
grid worker nodes. Examples of the latter are the so-called Pilot Agents, part
of the Workload Management System explained in the following section. All DIRAC
agents repeat the same logic in each iteration cycle: they monitor the service states,
 and respond by initiating actions (like job
submission or data transfer) which update the states of various system
entities.
\begin{figure}[tp!]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}
\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

@@

Fig.~\ref{fig:CacheReplicas-SM} depicts the state machine of the CacheReplicas entitiy. The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' replica entries, checks whether they are registered in a Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the state of the \textit{CacheReplicas} and the related \textit{Tasks}
entries to ``Failed''. Non-problematic files are updated to a
``Waiting'' state. The \textit{Stage Request Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by Storage Element (SE) prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable, or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{Stage Monitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the \textit{CacheReplicas} and their
corresponding \textit{Tasks} entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{Request Finalization Agent}. The \textit{Tasks} which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated \textit{Tasks} for particular replicas, the respective
\textit{CacheReplicas} entries are also removed. The state machine of the \textit{CacheReplicas}
entity is shown in Fig.~\ref{fig:CacheReplicas-SM}.

@@

\subsubsection{Control Abstractions}
As already explained, all agents repeat the same logic in every subsequent iteration: first they
read some entries of interest from the service database, then they process the cached data, and finally
they may write back or update entries, based on decisions from the processing step.
They can be naturally modeled as \textit{recursive processes}. Each process represents a distinct agent,
specifying its expected behavior when interacting with the service database.
Take as an example the code snippet
in Listing~\ref{listing2}, from the \textit{Request Preparation Agent}. 
Although much of the code is omitted for clarity, the necessary parts for
illustrating the basic idea are kept.
The first highlighted statement is the selection of all "New'' CacheReplicas
entries. What follows is retrieval of their metadata from the LFC,
external to this subsystem. Subsequently, list and dictionary manipulations
are done to group the retrieved data depending on the outcome.
Two lists of replica IDs are built before the last step: one for the problematic catalog entries,
and one for the successful sanity checks.
Finally, the last two highlighted code segments update the states 
of the corresponding CacheReplicas to "Failed'' and ``Waiting'' respectively.

\definecolor{light-gray}{gray}{0.90}
\lstset{language=Python,tabsize=2,caption={\textit{RequestPreparationAgent.py} code excerpt},
  stringstyle=\ttfamily,
  %basicstyle=\small,
  frame=tb,
  breaklines=true,
  showstringspaces=false,
  boxpos=t
 }
\begin{lstlisting}[label=listing2,escapechar=!,basicstyle=\ttfamily\fontsize{8}{7}\ \selectfont]
  def prepareNewReplicas( self ):
    !\colorbox{light-gray}{res = self.getNewReplicas()}!
    if not res['Value']:
      gLogger.info('There were no New replicas found')
      return res
    [...]
    # Obtain the replicas from the FileCatalog
    res = self.__getFileReplicas( fileSizes.keys() )
    if not res['OK']:
      return res
    failed.update( res['Value']['Failed'] )
    terminal = res['Value']['ZeroReplicas']
    fileReplicas = res['Value']['Replicas']
    [...]
    replicaMetadata = []
    for lfn, requestedSEs in replicas.items():
      lfnReplicas = fileReplicas[lfn]
      for requestedSE, replicaID in requestedSEs.items():
        if not requestedSE in lfnReplicas.keys():
          terminalReplicaIDs[replicaID] = "LFN not registered at requested SE"
          replicas[lfn].pop(requestedSE)
        else:
          replicaMetadata.append((replicaID, lfnReplicas[requestedSE], fileSizes[lfn]))

@@

\begin{figure*}[tp!]

@@

@@

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[tp!]
\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}

@@

To overcome the problem of state space explosion, several
variants to explicit model checking have gained popularity. Runtime
verification~\cite{DBLP:conf/dagstuhl/ColinM04,Havelund:2008:RVC:1424465.1424470}
aims to increase confidence in the correctness, but does not make claims for
absence of defects in a system.
The target program is instrumented manually, or automatically by scanning
function calls and variable assignments, 
in order to capture and emit relevant events which are then monitored by a
special component during execution.
This component resembles our monitoring process automaton used to cope with
state space explosion during model checking 
of DIRAC's WMS system. While runtime verification is useful for
capturing incorrect behavior of the system implementation, 
code instrumentation is not always a feasible option, considering the increase
of memory consumption.

@@

Static analysis is, in a sense, an extreme form of automated deductive
verification.  In its original form, deductive verification technology
allowed for the use of logics such as Floyd-Hoare logic~\cite{Hoa:69} for
showing the correctness of a software or hardware system by constructing
proofs thereof, based on mathematical descriptions of the behavior of
the system.  These days, the proofs involved are often discharged using a
mechanical or an automated theorem prover. Moreover, more advanced logics
such as Separation Logic~\cite{Rey:02} are used. Deductive verification
is applied quite successfully on software~\cite{PMSJP:12,KAE+:10}
and hardware~\cite{AATH:10}, but the effort required for conducting the
verification can be prohibiting, and can compare poorly to model checking.

@@

Practitioners
of model checking have already built sufficiently mature tools
that can be utilized (almost) as blackboxes by more traditional
software developers, hiding the mathematical theories under
the hood. Although some domain-specific knowledge is always necessary,
the descriptive guidelines we have proposed for obtaining
an abstract model can be easily 
applied to other large-scale industrial systems of a similar nature.
It goes without saying that the analyst must be careful to faithfully
describe the system behavior in the abstract model. With this critical step in mind, formal methods
are an important addition to the assembly of software analysis and verification techniques for concurrent systems.
\section{Future work}
\label{sec:Section_6}
Besides discovering behavioral problems, the 
model gives opportunities to analyze the system performance in the future.
Performance analysis aims at predicting the behavior of a system with respect to dynamic properties, such as 
the expected throughput, response time, resource usage, number of requests per unit time, etc.
For this, the formal model must be enriched with timing and probability quantities attached 
to certain activities of interest.  Adding this kind of information to traditional process algebra models
means getting in the domain of stochastic process algebra. Random variables with exponential distribution are
typically associated with the designated actions, to express system transitions with duration, which could be estimates
of processing time or communication overhead. Tools based on stochastic process algebra formalisms can
then generate the underlying Markov Chain model representation and solve it to obtain steady-state probability distributions over the state space. These measures can provide a good base for assessing the system's efficiency and reliability. 
Furthermore, by varying the quantitative inputs, alternative designs can be evaluated before actual implementation
takes place. One such stochastic process algebra is PEPA (Performance Evaluation Process Algebra \cite{PEPA}), supported
by an Eclipse Plug-in for modeling and analysis of systems. 
PEPA models are annotated with information about the duration of activities and their relative probabilities of occurence.
